---
title: "Creating presentation - ready tables with gt and gtExtras"
author: "Elena Mijares Lopez and Ujwal Neethipudi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

## Introduction

In this workshop, we will explore how to create tables using the gt and gtExtras packages, utilizing the pizza place dataset. We will start with a basic table using base R to display pizza information, then enhance our visualization using the gt package to create a more polished and user-friendly table. Finally, we will upgrade our visualization further with the gtExtras package, adding even more features and styles to improve the overall presentation of our pizza menu.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(gt)
library(gtExtras)
library(svglite)
library(tidyverse)
```

## The Data

Let's look at our pizza sales dataset:

```{r}
# Load and assign the pizza dataset to an object
data("pizzaplace")
pizza_data <- pizzaplace
head(pizza_data)
```

## Basic Table (Before gt)

First, let's create a summary table using basic R:

```{r}
pizza_summary <- pizza_data %>%
  group_by(type, size) %>%
  summarize(
    total_sales = sum(price),
    num_orders = n(),
    avg_price = mean(price),
    .groups = 'drop'
  )

# Display as basic table
print(pizza_summary)
```

## Basic gt Features

### Step 1: Basic gt Table

Let's start with a simple gt table:

```{r}
pizza_summary %>%
  gt(
    groupname_col = "type"
  )
```

### Step 2: Adding Formatting

Now let's add currency and number formatting:

```{r}
pizza_summary %>%
  gt(
    groupname_col = "type"
  ) %>%
  fmt_currency(
    columns = c(total_sales, avg_price),
    currency = "USD"
  ) %>%
  fmt_number(
    columns = num_orders,
    decimals = 0
  )
```

### Step 3: Adding Headers and Labels

Let's improve the presentation with better headers:

```{r}
pizza_summary %>%
  gt(
    groupname_col = "type"
  ) %>%
  fmt_currency(
    columns = c(total_sales, avg_price),
    currency = "USD"
  ) %>%
  fmt_number(
    columns = num_orders,
    decimals = 0
  ) %>%
  tab_header(
    title = "Pizza Sales Analysis",
    subtitle = "Sales breakdown by type and size"
  ) %>%
  cols_label(
    total_sales = "Total Sales",
    num_orders = "Number of Orders",
    avg_price = "Average Price",
    size = "Pizza Size"
  )
```

### Step 4: Adding Visual Elements with gtExtras

Now let's add some visual enhancements:

```{r}
pizza_summary %>%
  gt(
    groupname_col = "type"
  ) %>%
  fmt_currency(
    columns = c(total_sales, avg_price),
    currency = "USD"
  ) %>%
  fmt_number(
    columns = num_orders,
    decimals = 0
  ) %>%
  tab_header(
    title = "Pizza Sales Analysis",
    subtitle = "Sales breakdown by type and size"
  ) %>%
  cols_label(
    total_sales = "Total Sales",
    num_orders = "Number of Orders",
    avg_price = "Average Price",
    size = "Pizza Size"
  ) %>%
  gt_theme_538() %>%
  data_color(
    columns = c(total_sales),
    colors = scales::col_numeric(
      palette = c("white", "#1F77B4"),
      domain = NULL
    )
  )
```

## Advanced Features

### Sales Trends Visualization

Let's create a different view focusing on pizza types:

```{r}
pizza_data %>%
  # First calculate the total number of orders
  mutate(total_orders = n()) %>%
  group_by(type) %>%
  summarize(
    total_sales = sum(price),
    num_orders = n(),
    avg_price = mean(price),
    pct_total = num_orders/first(total_orders) * 100  # Fixed percentage calculation
  ) %>%
  ungroup() %>%
  gt() %>%
  # Basic formatting
  fmt_currency(columns = c(total_sales, avg_price), currency = "USD") %>%
  fmt_number(columns = num_orders, decimals = 0) %>%
  fmt_percent(columns = pct_total, decimals = 1) %>%
  
  # Enhanced styling
  gt_theme_nytimes() %>%
  
  # Add mini charts
  gt_plt_bar_pct(column = pct_total, scaled = TRUE) %>%
  
  # Enhanced header
  tab_header(
    title = md("**ðŸ• Pizza Sales Performance Analysis**"),
    subtitle = md("*Distribution of sales across pizza types*")
  ) %>%
  
  # Column formatting
  cols_width(
    type ~ px(150),
    total_sales ~ px(200)
  ) %>%
  
  # Conditional formatting
  tab_style(
    style = list(
      cell_fill(color = "#e8f4f8"),
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      columns = num_orders,
      rows = num_orders > mean(num_orders)
    )
  ) %>%
  
  # Color gradient for total sales
  data_color(
    columns = total_sales,
    colors = scales::col_numeric(
      palette = c("white", "steelblue"),
      domain = NULL
    )
  ) %>%
  
  # Footer elements
  tab_footnote(
    footnote = "Higher than average sales",
    locations = cells_body(
      columns = total_sales,
      rows = total_sales > mean(total_sales)
    )
  ) %>%
  tab_source_note(
    source_note = md("*Data source: Pizza Place Sales*")
  ) %>%
  
  # Table options
  tab_options(
    heading.title.font.size = px(24),
    heading.subtitle.font.size = px(16),
    heading.background.color = "#f6f6f6",
    column_labels.background.color = "#e6e6e6",
    table.border.top.width = px(3),
    table.border.top.color = "#4A4A4A",
    column_labels.font.weight = "bold",
    data_row.padding = px(12),
    source_notes.font.size = px(10)
  )
```

### Dark Theme Example

```{r}
pizza_data %>%
  group_by(type) %>%
  summarize(
    total_sales = sum(price),
    num_orders = n(),
    avg_price = mean(price),
    sales_trend = list(tapply(price, date, mean))
  ) %>%
  gt() %>%
  
  gt_theme_dark() %>%
  
  # Format numbers
  fmt_currency(
    columns = c(total_sales, avg_price),
    currency = "USD",
    placement = "right",
    decimals = 2,
    use_subunits = TRUE
  ) %>%
  fmt_number(
    columns = num_orders,
    decimals = 0,
    use_seps = TRUE
  ) %>%
  
  # Sparkline and bar plots
  gt_plt_sparkline(sales_trend, label = FALSE) %>%
  
  # gt_plt_bar(column = total_sales, color = "steelblue", width = 65) %>%
  
  # Header styling
  tab_header(
    title = md("ðŸŒ™ Night Mode Sales Dashboard"),
    subtitle = md("*Visualization of pizza sales*")
  ) %>%
  
  # Column labels
  cols_label(
    type = md("**Pizza Type**"),
    total_sales = md("**Total Revenue**"),
    num_orders = md("**Orders**"),
    avg_price = md("**Avg. Price**"),
    sales_trend = md("**Sales Trend**")
  ) %>%
  
  # Column alignment
  cols_align(
    align = "left",
    columns = type
  ) %>%
  cols_align(
    align = "center",
    columns = c(num_orders, sales_trend)
  ) %>%
  cols_align(
    align = "right",
    columns = c(total_sales, avg_price)
  ) %>%
  
  # Column width adjustments
  cols_width(
    type ~ px(150),
    total_sales ~ px(180),
    num_orders ~ px(120),
    avg_price ~ px(120),
    sales_trend ~ px(150)
  ) %>%
  
  # Add borders and styling
  tab_style(
    style = list(
      cell_borders(
        sides = c("left", "right"),
        color = "#404040",
        weight = px(1)
      )
    ),
    locations = cells_body()
  ) %>%
  
  # Highlight highest values
  tab_style(
    style = list(
      cell_fill(color = "#2d4a54"),
      cell_text(weight = "bold", color = "white")  # Ensure text is visible
    ),
    locations = cells_body(
      columns = total_sales,
      rows = total_sales == max(total_sales)
    )
  ) %>%
  
  # Add row striping
  opt_row_striping() %>%
  
  # Table styling
  tab_options(
    table.background.color = "#1a1a1a",
    heading.title.font.size = px(28),
    heading.subtitle.font.size = px(18),
    heading.padding = px(20),
    column_labels.background.color = "#2d2d2d",
    column_labels.font.weight = "bold",
    column_labels.padding = px(15),
    table.border.top.style = "solid",
    table.border.top.width = px(3),
    table.border.top.color = "white",
    table.border.bottom.color = "white",
    table.border.bottom.width = px(3),
    column_labels.border.bottom.color = "white",
    column_labels.border.bottom.width = px(2),
    data_row.padding = px(12),
    row.striping.background_color = "#2a2a2a",
    table.font.color = "white",
    table.font.size = px(14)
  ) %>%
  
  # Add source note
  tab_source_note(
    source_note = md("*Data source: Pizza Place Sales Dataset*")
  )
```

### Summary Rows Feature

```{r}
pizza_data %>%
  group_by(type) %>%
  summarize(
    total_sales = sum(price),
    num_orders = n(),
    avg_price = mean(price)
  ) %>%
  ungroup() %>%
  gt(
    groupname_col = "type"
  ) %>%
  fmt_currency(
    columns = c(total_sales, avg_price),
    currency = "USD"
  ) %>%
  fmt_number(
    columns = num_orders,
    decimals = 0
  ) %>%
  grand_summary_rows(
    columns = total_sales,
    fns = list(
      "Total" = ~sum(.)
    ),
    fmt = ~ fmt_currency(., currency = "USD")  
  ) %>%
  grand_summary_rows(
    columns = num_orders,
    fns = list(
      "Total" = ~sum(.)
    ),
    fmt = ~ fmt_number(., decimals = 0)  
  )

```

## Final Demo

### A Pizza Pricing Table Using the gt and gtExtras

```{r}
# First create the pricing structure with features
pizza_pricing <- pizza_data %>%
  group_by(type, size) %>%
  summarize(
    price = mean(price),
    .groups = 'drop'
  ) %>%
  group_by(type) %>%
  summarize(
    basic = min(price),
    medium = median(price),
    large = max(price)
  ) %>%
  mutate(
    type_display = case_when(
      type == "chicken" ~ "ðŸ— Chicken Lovers",
      type == "classic" ~ "ðŸ• Classic Choice",
      type == "supreme" ~ "ðŸ‘‘ Supreme Selection",
      type == "veggie" ~ "ðŸ¥¬ Veggie Delight"
    ),
    features = case_when(
      type == "chicken" ~ "âœ“ Fresh chicken âœ“ Special herbs âœ“ Premium cheese",
      type == "classic" ~ "âœ“ Traditional sauce âœ“ Classic toppings âœ“ Family favorite",
      type == "supreme" ~ "âœ“ Premium toppings âœ“ Extra cheese âœ“ Special blend",
      type == "veggie" ~ "âœ“ Fresh vegetables âœ“ Light cheese âœ“ Herb-infused"
    )
  )

# Create the enhanced pricing table
pizza_pricing %>%
  select(type_display, features, basic, medium, large) %>%
  gt() %>%
  tab_header(
    title = md("ðŸŒŸ **Yummy Pizza Price Plans** ðŸŒŸ"),
    subtitle = md("*Explore our range of artisanal pizzas crafted just for you*")
  ) %>%
  
  cols_label(
    type_display = "",
    features = md("**Includes**"),
    basic = md("**Basic Plan**"),
    medium = md("**Premium Plan**"),
    large = md("**Deluxe Plan**")
  ) %>%
  
  fmt_currency(
    columns = c(basic, medium, large),
    currency = "USD"
  ) %>%
  
  # Style plan headers
  tab_style(
    style = list(
      cell_fill(color = "#4287f5"),
      cell_text(color = "white", weight = "bold", size = px(14)),
      cell_borders(sides = "all", color = "white", weight = px(2))
    ),
    locations = cells_column_labels(columns = basic)
  ) %>%
  
  tab_style(
    style = list(
      cell_fill(color = "#2ecc71"),
      cell_text(color = "white", weight = "bold", size = px(14)),
      cell_borders(sides = "all", color = "white", weight = px(2))
    ),
    locations = cells_column_labels(columns = medium)
  ) %>%
  
  tab_style(
    style = list(
      cell_fill(color = "#e74c3c"),
      cell_text(color = "white", weight = "bold", size = px(14)),
      cell_borders(sides = "all", color = "white", weight = px(2))
    ),
    locations = cells_column_labels(columns = large)
  ) %>%
  
  # Style pizza types
  tab_style(
    style = list(
      cell_text(weight = "bold", size = px(14)),
      cell_fill(color = "#f8f9fa"),
      cell_borders(sides = "all", color = "#dee2e6", weight = px(1))
    ),
    locations = cells_body(columns = type_display)
  ) %>%
  
  # Add borders
  tab_style(
    style = list(
      cell_borders(sides = "all", color = "#dee2e6", weight = px(2))
    ),
    locations = cells_body()
  ) %>%
  
  # Alignment
  cols_align(
    align = "center",
    columns = c(basic, medium, large)
  ) %>%
  
  cols_align(
    align = "left",
    columns = c(type_display, features)
  ) %>%
  
  # Column widths
  cols_width(
    type_display ~ px(180),
    features ~ px(250),
    everything() ~ px(150)
  ) %>%
  
  # Add plan features spanner
  tab_spanner(
    label = md("**ðŸŽ¯ Select Your Perfect Plan ðŸŽ¯**"),
    columns = c(basic, medium, large)
  ) %>%
  
  # Style the prices
  tab_style(
    style = list(
      cell_text(weight = "bold", size = px(16)),
      cell_fill(color = "#ffffff")
    ),
    locations = cells_body(
      columns = c(basic, medium, large)
    )
  ) %>%
  
  # Add row striping
  opt_row_striping(row_striping = TRUE) %>%
  
  # Table styling
  tab_options(
    heading.background.color = "#f8f9fa",
    heading.title.font.size = px(28),
    heading.subtitle.font.size = px(16),
    heading.padding = px(20),
    column_labels.background.color = "#f8f9fa",
    table.border.top.width = px(3),
    table.border.bottom.width = px(3),
    table.border.top.color = "#4e73df",
    table.border.bottom.color = "#4e73df",
    data_row.padding = px(15),
    row.striping.background_color = "#f9f9f9",
    heading.title.font.weight = "bold"
  ) %>%
  
  # Add features styling
  tab_style(
    style = list(
      cell_borders(sides = "all", color = "#dee2e6", weight = px(1)),
      cell_fill(color = "#e8f4f8")
    ),
    locations = cells_body(columns = features)
  ) %>%
  
  # Add badges/footnotes
  tab_footnote(
    footnote = md("ðŸ”¥ **Most Popular Choice**"),
    locations = cells_column_labels(columns = medium)
  ) %>%
  
  tab_footnote(
    footnote = md("â­ **Best Value**"),
    locations = cells_column_labels(columns = large)
  ) %>%
  
  # Source notes
  tab_source_note(
    source_note = md(
      "âœ¨ *All plans include our signature sauce and premium cheese*  |  
      ðŸ›µ *Free delivery on orders over $30*  |  
      ðŸ’« *Customization available for all plans*"
    )
  )
```

